The algebraic hierarchy in type theory using type classes

### 2010-02-25 processed.tex
first commit

\documentclass[a4paper,10pt, runningheads]{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage{ucs}
% \usepackage[utf8x,utf8]{inputenc}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
%include coq2tex.tex
\arrayhs%i.e. no page breaks in code blocks
\divide\abovedisplayskip 4
\divide\belowdisplayskip 4
\begin{document}
\title{The algebraic hierarchy in type theory using type classes}
\author{Bas Spitters \and Eelis van der Weegen}
\institute{Radboud University Nijmegen}
\maketitle
\begin{abstract}
We propose generic design patterns to organize algebraic structures using type classes inside the
Coq system. We support multiple inheritance, sharing of notations and theories, and automated
structure inference. A library for many-sorted universal algebra abstracts from the specific
algebraic structures and aids the organization of the library. Category theory provides a further
useful abstraction. We provide abstract
interfaces for $\N$, $\Z$, $\Q$ as the initial semiring, the initial ring and a field of
fractions. Similarly, we provide an interface for polynomials over a ring $R$ using $R$-algebras.
We show that the standard implementations of these are in fact instances of these type classes.
We report on the efficiency penality of these abstractions.

@@

Several solutions have been proposed for the Coq theorem prover~\cite{Coq,BC04}: dependent
records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a.\ telescopes), packed classes~\cite{Packed} and
occasionally modules. Although the latter option is quickly discarded%
\footnote{Modules are not first class, but currently they seem to be superior for extraction to
OCaml~\cite{letouzey02}. Modules may also be seen as parametrised sections. They allow management
of namespaces, notations, hints, etc. An |Import| command imports all the constants from a
module. Thus allowing the user to type |B| instead of |A.B|. Similar functionality would be useful
for type classes.}
algebraic theories as `interfaces', as opposed to `modules', has been an important
guiding principle~\cite{Packed}.

@@

\section{Preliminaries}
\subsection{Coq type theory}
The Coq proof assistant is based on the calculus of inductive
constructions~\cite{CoquandHuet,CoquandPaulin}.\marginpar{More}
This type theory lacks quotient types, as it would make type checking undecidable. Instead it
supports setoids (aka Bishop sets), a pair of a type and an equivalence
relation~\cite{Bishop67,Hofmann,Capretta}. Recently, improved support has been added to the
system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called |Proper|.
As we will see in Section~\ref{univ2} below this functionality can also be used to work with
quotient algebras.

@@

How should we define reflexive, transitive, and symmetric relations? In a first attempt, we
might define:
%format → = "\rightarrow"
%format Π = "\Pi"
\begin{code}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: rr_carrier → rr_carrier → Prop
    ; rr_proof: Π x, rr_rel x x }.
\end{code}
and similar for transitive and symmetric relations.

With this approach, we run into problems when we try to define |EquivalenceRelation| in terms of
|ReflexiveRelation| and its two siblings, because when posit three objects of type
|ReflexiveRelation|, |TransitiveRelation|, and |SymmetricRelation|, respectively, we have to
add equalities expressing that they talk about the same carrier and relation, and these equalities
are a nightmare to work with (because the three carriers and relations won't be convertible).
Technically, this leads to the research in manifest fields, the \emph{with} construction, to which
we return in Section~\ref{manifest}.

As a first step we would define |ReflexiveRelation| as:
\begin{code}
  Record ReflexiveRelation (carrier: Type): Type :=
    { rr_rel: rr_carrier → rr_carrier → Prop
    ; rr_proof: Π x, rr_rel x x }.
\end{code}

This would solve the problem for the carrier, but |EquivalenceRelation| would still need equalities
for the relation:
\begin{code}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_refl: ReflexiveRelation er_carrier
    ; er_sym: SymmetricRelation er_carrier
    ; er_trans: TransitiveRelation er_trans
    ; same_rel: rr_rel er_refl = sr_rel er_sym
    ; same_rel': rr_rel er_refl = tr_rel er_trans }.
\end{code}

A natural solution is to pull the relation out of |ReflexiveRelation| (and its siblings) as well:
\begin{code}
  Record ReflexiveRelation (carrier: Type)
                           (rel: carrier → carrier → Prop): Prop :=
    { rr_proof: Π x, rr_rel x x }.
\end{code}
or equivalently:
\begin{code}
  Definition ReflexiveRelation (carrier: Type)
                               (rel: carrier → carrier → Prop): Prop :=
    Π x, rr_rel x x.
\end{code}
This is how reflexivity is defined in the standard library, and this is entirely sensible. The
change from |Type| to |Prop| reflects the fact that |ReflexiveRelation| is now strictly a property,
proofs of which can be nicely kept opaque;%
\footnote{Coq distinguishes between transparent and opague definitions. The former can be unfolded,
whereas this is impossible for the latter.}
we will return to this issue in Section~\ref{explicit}.
We can now define |EquivalenceRelation| without awkward equalities:
\begin{code}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_rel: er_carrier → er_carrier → Prop
    ; er_refl: ReflexiveRelation er_carrier er_rel
    ; er_sym: SymmetricRelation er_carrier er_rel
    ; er_trans: TransitiveRelation er_trans er_rel }.
\end{code}

But looking at |EquivalenceRelation|, we see that it now essentially has the same form as the
original naive |ReflexiveRelation| record we started with. Clearly, equivalence relations
and reflexive relations are analogous concepts, and by the same argument we
conclude that |EquivalenceRelation| should really be defined as:

\begin{code}
Record EquivalenceRelation (er_carrier: Type)
                             (er_rel: er_carrier → er_carrier → Prop): Prop :=
  { er_refl: ReflexiveRelation er_carrier er_rel
  ; er_sym: SymmetricRelation er_carrier er_rel
  ; er_trans: TransitiveRelation er_trans er_rel }.
\end{code}

After all, when we define a type with
two different equivalence relations on it,
such as the quotient of a setoid,
we have the same considerations as for |ReflexiveRelation| above.

At the risk of overstressing the point, let us look at how the same problem and solution keep
appearing in composite structures. Suppose we use a definition of |Monoid| that bundles the
relation. Then if we proceed to define |SemiRing| as a double |Monoid|, we face a dilemma:
\begin{code}
  Record SemiRing A: Type :=
    { mult_monoid: Monoid A 
    ; plus_monoid: Monoid A 
    ; plus_comm: Commutative (mon_op plus_monoid)
    ; mult_comm: Commutative (mon_op mult_monoid)
    ; mult_0_l: Π x: A, ?
\end{code}
Which equality do we use in the expression ?, |mult_monoid|'s equality or |plus_monoid|'s?
Again, for this definition to be usable at all, we would need to add an equality expressing that the
two monoids contain the same equivalence relation. And again, the proper solution is to change
Monoid and unbundle the relation. But here, we have a choice as to how much we unbundle exactly.
(For the moment we ignore the question of whether the unit and binary operator should be unbundled
as well, focusing only on the equivalence relation and its equivalence proof.)

We \emph{could} become over-eager and decide to unbundle the equivalence proof term as well:
\begin{code}
Record Monoid (A: Type) (R: A → A → Prop)
  (e: Equivalence R) : Type := ...
\end{code}
However, this way if we happen to have two proofs |p| and |q| that a certain relation is an
equivalence, we end up with two non-convertible ways to express that the relation (together with
some unit and binary operation) form a Monoid: 'Monoid A R p ...' and 'Monoid A R q ...'. We must
either try to deal with this (by proving that Monoid proofs with different Equivalence proofs are
interchangable, and hope for automation, such as type class instance resolution, to use them
interchangably), or take great care to ensure that |Equivalence| proof terms are always shared,
essentially elevating them to `values', which goes against the principle of proof
irrelevance.\marginpar{What if we add proof irrelevance?}

Neither of these options are very appealing. We really should just make |Monoid| take as parameters
only the operations/relations. If we treat the monoid's unit and binary operation the same way, this
means that |Monoid| ends up in |Prop|, and that its instances can be left opaque, which is really
how things ought to be.

>>>>

\documentclass[a4paper,10pt, runningheads]{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage{ucs}
% \usepackage[utf8x,utf8]{inputenc}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
%% ODER: format ==         = "\mathrel{==}"
%% ODER: format /=         = "\neq "
%
%
\makeatletter
\@ifundefined{lhs2tex.lhs2tex.sty.read}%
  {\@namedef{lhs2tex.lhs2tex.sty.read}{}%
   \newcommand\SkipToFmtEnd{}%
   \newcommand\EndFmtInput{}%
   \long\def\SkipToFmtEnd#1\EndFmtInput{}%
  }\SkipToFmtEnd

\newcommand\ReadOnlyOnce[1]{\@ifundefined{#1}{\@namedef{#1}{}}\SkipToFmtEnd}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\DeclareFontFamily{OT1}{cmtex}{}
\DeclareFontShape{OT1}{cmtex}{m}{n}
  {<5><6><7><8>cmtex8
   <9>cmtex9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmtex10}{}
\DeclareFontShape{OT1}{cmtex}{m}{it}
  {<-> ssub * cmtt/m/it}{}
\newcommand{\texfamily}{\fontfamily{cmtex}\selectfont}
\DeclareFontShape{OT1}{cmtt}{bx}{n}
  {<5><6><7><8>cmtt8
   <9>cmbtt9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmbtt10}{}
\DeclareFontShape{OT1}{cmtex}{bx}{n}
  {<-> ssub * cmtt/bx/n}{}
\newcommand{\tex}[1]{\text{\texfamily#1}}	% NEU

\newcommand{\Sp}{\hskip.33334em\relax}

\newcommand{\Conid}[1]{\mathit{#1}}
\newcommand{\Varid}[1]{\mathit{#1}}
\newcommand{\anonymous}{\kern0.06em \vbox{\hrule\@width.5em}}
\newcommand{\plus}{\mathbin{+\!\!\!+}}
\newcommand{\bind}{\mathbin{>\!\!\!>\mkern-6.7mu=}}
\newcommand{\rbind}{\mathbin{=\mkern-6.7mu<\!\!\!<}}% suggested by Neil Mitchell
\newcommand{\sequ}{\mathbin{>\!\!\!>}}
\renewcommand{\leq}{\leqslant}
\renewcommand{\geq}{\geqslant}
\usepackage{polytable}

%mathindent has to be defined
\@ifundefined{mathindent}%
  {\newdimen\mathindent\mathindent\leftmargini}%
  {}%

\def\resethooks{%
  \global\let\SaveRestoreHook\empty
  \global\let\ColumnHook\empty}
\newcommand*{\savecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\savecolumns[#1]}}
\newcommand*{\restorecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\restorecolumns[#1]}}
\newcommand*{\aligncolumn}[2]%
  {\g@addto@macro\ColumnHook{\column{#1}{#2}}}

\resethooks

\newcommand{\onelinecommentchars}{\quad-{}- }
\newcommand{\commentbeginchars}{\enskip\{-}
\newcommand{\commentendchars}{-\}\enskip}

\newcommand{\visiblecomments}{%
  \let\onelinecomment=\onelinecommentchars
  \let\commentbegin=\commentbeginchars
  \let\commentend=\commentendchars}

\newcommand{\invisiblecomments}{%
  \let\onelinecomment=\empty
  \let\commentbegin=\empty
  \let\commentend=\empty}

\visiblecomments

\newlength{\blanklineskip}
\setlength{\blanklineskip}{0.66084ex}

\newcommand{\hsindent}[1]{\quad}% default is fixed indentation
\let\hspre\empty
\let\hspost\empty
\newcommand{\NB}{\textbf{NB}}
\newcommand{\Todo}[1]{$\langle$\textbf{To do:}~#1$\rangle$}

\EndFmtInput
\makeatother
%
%
%
%
%
%
% This package provides two environments suitable to take the place
% of hscode, called "plainhscode" and "arrayhscode". 
%
% The plain environment surrounds each code block by vertical space,
% and it uses \abovedisplayskip and \belowdisplayskip to get spacing
% similar to formulas. Note that if these dimensions are changed,
% the spacing around displayed math formulas changes as well.
% All code is indented using \leftskip.
%
% Changed 19.08.2004 to reflect changes in colorcode. Should work with
% CodeGroup.sty.
%
\ReadOnlyOnce{polycode.fmt}%
\makeatletter

\newcommand{\hsnewpar}[1]%
  {{\parskip=0pt\parindent=0pt\par\vskip #1\noindent}}

% can be used, for instance, to redefine the code size, by setting the
% command to \small or something alike
\newcommand{\hscodestyle}{}

% The command \sethscode can be used to switch the code formatting
% behaviour by mapping the hscode environment in the subst directive
% to a new LaTeX environment.

\newcommand{\sethscode}[1]%
  {\expandafter\let\expandafter\hscode\csname #1\endcsname
   \expandafter\let\expandafter\endhscode\csname end#1\endcsname}

% "compatibility" mode restores the non-polycode.fmt layout.

\newenvironment{compathscode}%
  {\par\noindent
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \par\noindent
   \ignorespacesafterend}

\newcommand{\compaths}{\sethscode{compathscode}}

% "plain" mode is the proposed default.
% It should now work with \centering.
% This required some changes. The old version
% is still available for reference as oldplainhscode.

\newenvironment{plainhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\hspre\(\let\hspost\)%
   \pboxed}%
  {\endpboxed%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

\newenvironment{oldplainhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

% Here, we make plainhscode the default environment.

\newcommand{\plainhs}{\sethscode{plainhscode}}
\newcommand{\oldplainhs}{\sethscode{oldplainhscode}}
\plainhs

% The arrayhscode is like plain, but makes use of polytable's
% parray environment which disallows page breaks in code blocks.

\newenvironment{arrayhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\parray}%
  {\endparray\)%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

\newcommand{\arrayhs}{\sethscode{arrayhscode}}

% The mathhscode environment also makes use of polytable's parray 
% environment. It is supposed to be used only inside math mode 
% (I used it to typeset the type rules in my thesis).

\newenvironment{mathhscode}%
  {\parray}{\endparray}

\newcommand{\mathhs}{\sethscode{mathhscode}}

% texths is similar to mathhs, but works in text mode.

\newenvironment{texthscode}%
  {\(\parray}{\endparray\)}

\newcommand{\texths}{\sethscode{texthscode}}

% The framed environment places code in a framed box.

\def\codeframewidth{\arrayrulewidth}
\RequirePackage{calc}

\newenvironment{framedhscode}%
  {\parskip=\abovedisplayskip\par\noindent
   \hscodestyle
   \arrayrulewidth=\codeframewidth
   \tabular{@{}|p{\linewidth-2\arraycolsep-2\arrayrulewidth-2pt}|@{}}%
   \hline\framedhslinecorrect\\{-1.5ex}%
   \let\endoflinesave=\\
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \framedhslinecorrect\endoflinesave{.5ex}\hline
   \endtabular
   \parskip=\belowdisplayskip\par\noindent
   \ignorespacesafterend}

\newcommand{\framedhslinecorrect}[2]%
  {#1[#2]}

\newcommand{\framedhs}{\sethscode{framedhscode}}

% The inlinehscode environment is an experimental environment
% that can be used to typeset displayed code inline.

\newenvironment{inlinehscode}%
  {\(\def\column##1##2{}%
   \let\>\undefined\let\<\undefined\let\\\undefined
   \newcommand\>[1][]{}\newcommand\<[1][]{}\newcommand\\[1][]{}%
   \def\fromto##1##2##3{##3}%
   \def\nextline{}}{\) }%

\newcommand{\inlinehs}{\sethscode{inlinehscode}}

% The joincode environment is a separate environment that
% can be used to surround and thereby connect multiple code
% blocks.

\newenvironment{joincode}%
  {\let\orighscode=\hscode
   \let\origendhscode=\endhscode
   \def\endhscode{\def\hscode{\endgroup\def\@currenvir{hscode}\\}\begingroup}
   %\let\SaveRestoreHook=\empty
   %\let\ColumnHook=\empty
   %\let\resethooks=\empty
   \orighscode\def\hscode{\endgroup\def\@currenvir{hscode}}}%
  {\origendhscode
   \global\let\hscode=\orighscode
   \global\let\endhscode=\origendhscode}%

\makeatother
\EndFmtInput
%

  % Without these, lhs2tex uses three-line equal signs.

\arrayhs%i.e. no page breaks in code blocks
\divide\abovedisplayskip 4
\divide\belowdisplayskip 4
\begin{document}
\title{The algebraic hierarchy in type theory using type classes}
\author{Bas Spitters \and Eelis van der Weegen}
\institute{Radboud University Nijmegen}
\maketitle
\begin{abstract}
We propose generic design patterns to organize algebraic structures using type classes inside the
Coq system. We support multiple inheritance, sharing of notations and theories, and automated
structure inference. A library for many-sorted universal algebra abstracts from the specific
algebraic structures and aids the organization of the library. Category theory provides a further
useful abstraction. We provide abstract
interfaces for $\N$, $\Z$, $\Q$ as the initial semiring, the initial ring and a field of
fractions. Similarly, we provide an interface for polynomials over a ring $R$ using $R$-algebras.
We show that the standard implementations of these are in fact instances of these type classes.
We report on the efficiency penality of these abstractions.

@@

Several solutions have been proposed for the Coq theorem prover~\cite{Coq,BC04}: dependent
records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a.\ telescopes), packed classes~\cite{Packed} and
occasionally modules. Although the latter option is quickly discarded%
\footnote{Modules are not first class, but currently they seem to be superior for extraction to
OCaml~\cite{letouzey02}. Modules may also be seen as parametrised sections. They allow management
of namespaces, notations, hints, etc. An \ensuremath{\Conid{Import}} command imports all the constants from a
module. Thus allowing the user to type \ensuremath{\Conid{B}} instead of \ensuremath{\Conid{A.B}}. Similar functionality would be useful
for type classes.}
algebraic theories as `interfaces', as opposed to `modules', has been an important
guiding principle~\cite{Packed}.

@@

\section{Preliminaries}
\subsection{Coq type theory}
The Coq proof assistant is based on the calculus of inductive
constructions~\cite{CoquandHuet,CoquandPaulin}.\marginpar{More}
This type theory lacks quotient types, as it would make type checking undecidable. Instead it
supports setoids (aka Bishop sets), a pair of a type and an equivalence
relation~\cite{Bishop67,Hofmann,Capretta}. Recently, improved support has been added to the
system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called \ensuremath{\Conid{Proper}}.
As we will see in Section~\ref{univ2} below this functionality can also be used to work with
quotient algebras.

@@

How should we define reflexive, transitive, and symmetric relations? In a first attempt, we
might define:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathbf{Record}\;\Conid{ReflexiveRelation:}\;\Conid{Type}\;\Conid{:=}\;{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\hspace*{0mm}\{\Varid{rr\char95 carrier:}\;\Conid{Type}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{};\Varid{rr\char95 rel:}\;\Varid{rr\char95 carrier}\;\rightarrow\;\Varid{rr\char95 carrier}\;\rightarrow\;\Conid{Prop}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{};\Varid{rr\char95 proof:}\;\Pi\;{}\<[20]%
\>[20]{}\Varid{x},\Varid{rr\char95 rel}\;\Varid{x}\;\Varid{x}\mskip1.5mu\}\;.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
and similar for transitive and symmetric relations.

With this approach, we run into problems when we try to define \ensuremath{\Conid{EquivalenceRelation}} in terms of
\ensuremath{\Conid{ReflexiveRelation}} and its two siblings, because when posit three objects of type
\ensuremath{\Conid{ReflexiveRelation}}, \ensuremath{\Conid{TransitiveRelation}}, and \ensuremath{\Conid{SymmetricRelation}}, respectively, we have to
add equalities expressing that they talk about the same carrier and relation, and these equalities
are a nightmare to work with (because the three carriers and relations won't be convertible).
Technically, this leads to the research in manifest fields, the \emph{with} construction, to which
we return in Section~\ref{manifest}.

As a first step we would define \ensuremath{\Conid{ReflexiveRelation}} as:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathbf{Record}\;\Conid{ReflexiveRelation}\;(\Varid{carrier:}\;\Conid{Type})\;\mathbin{:}\;\Conid{Type}\;\Conid{:=}\;{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\hspace*{0mm}\{\Varid{rr\char95 rel:}\;\Varid{rr\char95 carrier}\;\rightarrow\;\Varid{rr\char95 carrier}\;\rightarrow\;\Conid{Prop}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{};\Varid{rr\char95 proof:}\;\Pi\;{}\<[20]%
\>[20]{}\Varid{x},\Varid{rr\char95 rel}\;\Varid{x}\;\Varid{x}\mskip1.5mu\}\;.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

This would solve the problem for the carrier, but \ensuremath{\Conid{EquivalenceRelation}} would still need equalities
for the relation:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathbf{Record}\;\Conid{EquivalenceRelation:}\;\Conid{Type}\;\Conid{:=}\;{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\hspace*{0mm}\{\Varid{er\char95 carrier:}\;\Conid{Type}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{};\Varid{er\char95 refl:}\;\Conid{ReflexiveRelation}\;\Varid{er\char95 carrier}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{};\Varid{er\char95 sym:}\;\Conid{SymmetricRelation}\;\Varid{er\char95 carrier}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{};\Varid{er\char95 trans:}\;\Conid{TransitiveRelation}\;\Varid{er\char95 trans}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{};\Varid{same\char95 rel:}\;\Varid{rr\char95 rel}\;\Varid{er\char95 refl}\;\mathrel{=}\;\Varid{sr\char95 rel}\;\Varid{er\char95 sym}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{};\Varid{same\char95 rel':}\;\Varid{rr\char95 rel}\;\Varid{er\char95 refl}\;\mathrel{=}\;\Varid{tr\char95 rel}\;\Varid{er\char95 trans}\mskip1.5mu\}\;.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

A natural solution is to pull the relation out of \ensuremath{\Conid{ReflexiveRelation}} (and its siblings) as well:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
\column{28}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathbf{Record}\;\Conid{ReflexiveRelation}\;(\Varid{carrier:}\;\Conid{Type})\;{}\<[E]%
\\
\>[3]{}\hsindent{25}{}\<[28]%
\>[28]{}(\Varid{rel:}\;\Varid{carrier}\;\rightarrow\;\Varid{carrier}\;\rightarrow\;\Conid{Prop})\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}\;{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\hspace*{0mm}\{\Varid{rr\char95 proof:}\;\Pi\;{}\<[20]%
\>[20]{}\Varid{x},\Varid{rr\char95 rel}\;\Varid{x}\;\Varid{x}\mskip1.5mu\}\;.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
or equivalently:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
\column{32}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathbf{Definition}\;\Conid{ReflexiveRelation}\;(\Varid{carrier:}\;\Conid{Type})\;{}\<[E]%
\\
\>[3]{}\hsindent{29}{}\<[32]%
\>[32]{}(\Varid{rel:}\;\Varid{carrier}\;\rightarrow\;\Varid{carrier}\;\rightarrow\;\Conid{Prop})\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}\;{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Pi\;{}\<[8]%
\>[8]{}\Varid{x},\Varid{rr\char95 rel}\;\Varid{x}\;\Varid{x.}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
This is how reflexivity is defined in the standard library, and this is entirely sensible. The
change from \ensuremath{\Conid{Type}} to \ensuremath{\Conid{Prop}} reflects the fact that \ensuremath{\Conid{ReflexiveRelation}} is now strictly a property,
proofs of which can be nicely kept opaque;%
\footnote{Coq distinguishes between transparent and opague definitions. The former can be unfolded,
whereas this is impossible for the latter.}
we will return to this issue in Section~\ref{explicit}.
We can now define \ensuremath{\Conid{EquivalenceRelation}} without awkward equalities:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathbf{Record}\;\Conid{EquivalenceRelation:}\;\Conid{Type}\;\Conid{:=}\;{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\hspace*{0mm}\{\Varid{er\char95 carrier:}\;\Conid{Type}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{};\Varid{er\char95 rel:}\;\Varid{er\char95 carrier}\;\rightarrow\;\Varid{er\char95 carrier}\;\rightarrow\;\Conid{Prop}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{};\Varid{er\char95 refl:}\;\Conid{ReflexiveRelation}\;\Varid{er\char95 carrier}\;\Varid{er\char95 rel}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{};\Varid{er\char95 sym:}\;\Conid{SymmetricRelation}\;\Varid{er\char95 carrier}\;\Varid{er\char95 rel}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{};\Varid{er\char95 trans:}\;\Conid{TransitiveRelation}\;\Varid{er\char95 trans}\;\Varid{er\char95 rel}\mskip1.5mu\}\;.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

But looking at \ensuremath{\Conid{EquivalenceRelation}}, we see that it now essentially has the same form as the
original naive \ensuremath{\Conid{ReflexiveRelation}} record we started with. Clearly, equivalence relations
and reflexive relations are analogous concepts, and by the same argument we
conclude that \ensuremath{\Conid{EquivalenceRelation}} should really be defined as:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{Record}\;\Conid{EquivalenceRelation}\;(\Varid{er\char95 carrier:}\;\Conid{Type})\;{}\<[E]%
\\
\>[B]{}\hsindent{30}{}\<[30]%
\>[30]{}(\Varid{er\char95 rel:}\;\Varid{er\char95 carrier}\;\rightarrow\;\Varid{er\char95 carrier}\;\rightarrow\;\Conid{Prop})\;\mathbin{:}\;\Conid{Prop}\;\Conid{:=}\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\hspace*{0mm}\{\Varid{er\char95 refl:}\;\Conid{ReflexiveRelation}\;\Varid{er\char95 carrier}\;\Varid{er\char95 rel}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{};\Varid{er\char95 sym:}\;\Conid{SymmetricRelation}\;\Varid{er\char95 carrier}\;\Varid{er\char95 rel}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{};\Varid{er\char95 trans:}\;\Conid{TransitiveRelation}\;\Varid{er\char95 trans}\;\Varid{er\char95 rel}\mskip1.5mu\}\;.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

After all, when we define a type with
two different equivalence relations on it,
such as the quotient of a setoid,
we have the same considerations as for \ensuremath{\Conid{ReflexiveRelation}} above.

At the risk of overstressing the point, let us look at how the same problem and solution keep
appearing in composite structures. Suppose we use a definition of \ensuremath{\Conid{Monoid}} that bundles the
relation. Then if we proceed to define \ensuremath{\Conid{SemiRing}} as a double \ensuremath{\Conid{Monoid}}, we face a dilemma:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathbf{Record}\;\Conid{SemiRing}\;\Conid{A:}\;\Conid{Type}\;\Conid{:=}\;{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\hspace*{0mm}\{\Varid{mult\char95 monoid:}\;\Conid{Monoid}\;\Conid{A}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{};\Varid{plus\char95 monoid:}\;\Conid{Monoid}\;\Conid{A}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{};\Varid{plus\char95 comm:}\;\Conid{Commutative}\;(\Varid{mon\char95 op}\;\Varid{plus\char95 monoid}){}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{};\Varid{mult\char95 comm:}\;\Conid{Commutative}\;(\Varid{mon\char95 op}\;\Varid{mult\char95 monoid}){}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{};\Varid{mult\char95 0\char95 l:}\;\Pi\;{}\<[20]%
\>[20]{}\Varid{x:}\;\Conid{A},\Varid{?}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Which equality do we use in the expression ?, \ensuremath{\Varid{mult\char95 monoid}}'s equality or \ensuremath{\Varid{plus\char95 monoid}}'s?
Again, for this definition to be usable at all, we would need to add an equality expressing that the
two monoids contain the same equivalence relation. And again, the proper solution is to change
Monoid and unbundle the relation. But here, we have a choice as to how much we unbundle exactly.
(For the moment we ignore the question of whether the unit and binary operator should be unbundled
as well, focusing only on the equivalence relation and its equivalence proof.)
